---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# likelihood.md.series.systems

`likelihood.md.series.systems` is an R package for estimating the parameters of
a series system model from masked data. It provides a flexible and intuitive
interface for specifying the model and performing maximum likelihood estimation (MLE).

This R package provides a set of functions for generating MLEs for the lifetime
parameters of the components in a series systems and other related characteristics
from data that *masks* the component cause of failure, and also the system lifetime.

Masked data comes in a variety of forms:

1. The system lifetime can be masked in three related ways.
   Right censoring occurs when the system under observation
   is only known to have survived for some minimum length of time.
   Left censoring occurs when the system under observation
   is only know to have survived for some maximum length of time.
   Finally, interval censoring occurs when the system under observation
   is only known to have survived between some minimum and maximum
   length of time.

   In the unmasked situation, we know precisely how long the system
   under observation survived.

2. Regardless of how the series system lifetime is masked, the
   lifetime of the components may be masked in any of the ways
   described in item (1). There is, additionally, another kind of
   masking we would like to consider. What if we do not observe
   any of the component lifetimes, and instead are only given
   a (potentially masked) series system lifetime, and a *candidate
   set* of component indexes which plausibly contains the failed
   component index.

   For a series system of $m$ components, the candidate sets are
   subsets of $\{1,\ldots,m\}$.

## Installation

You can install the development version from GitHub with:

```{r install, eval=FALSE}
# install.packages("devtools")
devtools::install_github("queelius/likelihood.md.series.systems")
```

## Usage

This package provides a comprehensive framework for maximum likelihood estimation (MLE) of series system parameters from masked data. It is based on [likelihood.model](https://queelius.github.io/likelihood.model), in particular [likelihood contributions](https://queelius.github.io/likelihood.model/articles/likelihood-contributions.html). We provide the likelihood 
contributions for a variety of observation types in this R package, and then use the 
`likelihood.model` package to work with the resultant likelihood model, e.g., for finding MLEs,
bootstrapping confidence intervals, and so on.

We also provide for data imputation, fake data, and so on. These are oftenbased on the
conditions the model assumes, and so are provided in this package.


# Usage

When we construct a likelihood contribution model, we so so by specifying
the contributions of each observation type. For example, if we have
a series system with three components, and we observe the system lifetime
and the lifetimes of the first two components, we would specify the
likelihood contributions as follows:

```{r model-example, eval = FALSE}
my_model <- likelihood_contr_model$new(
    obs_type = function(df) {
      ifelse(df$right_censoring,
        "exact_fail_time_with_cand_set_c1_c2_c3",
        "right_censored")
    },

    logliks = list(
        ...
    )
)
```

Now, we may call, for instance, `fit(my_model, data)` to fit the model.
We provide a default method for `fit` based on MLE, but you can also
provide your own method, e.g., based on a customized algorithm for efficiently
finding estimates for a particular type of series system for particular types
of data.

Regardless of the outcome, ideally it will return an `mle` object (from the
[algebraic.mle](https://queelius.github.io/algebraic.mle) package),
in which case a host of additional methods are available to you, such as
`predict`, `confint`, `sample`, etc. This object makes it easy to perform various
analyses on your fitted model.

```{r, eval = FALSE}
aic(fit)
bias(fit)
vcov(fit)
predict(fit, new_data)
confint(fit)
sample(fit, method = "asymptotic", n = 1000)
```

Note that the `likelihood.md.series.systems` package provides a number of
likelihood contributions for common observation types. These are described
in the package documentation. You can also provide your own likelihood
contributions, and we provide a number of functions to make this easier.

# Assumptions

Some of the models in this package make explicit assumptions about the data.
We provide various functions to help you check these assumptions, to
impute data that satisfies these assumptions, and to generate fake data
that satisfies these assumptions.

Sometimes, we generate fake data to create an implicit prior distribution
for the parameters.

# Model Selection

We provide some wrappers for model selection, such as AIC, BIC, and so on.
We also provide specializations and constraint functions for the parameters.

For example, we may have a strong belief that the components are Weibull distributed,
and furthermore, that they are more or less on the same scale with slight differences
in shape. In this case, we can simplify the model by assuming $\lambda_1 = \cdots = \lambda_m$
and then only fitting $\hat\lambda$ and $\hat k_1, \ldots, \hat k_m$, reducing a
$2m$ parameter model to a simpler $m+1$ parameter model. This is not an unrealistic
assumption, since if a series system is well-designed, then the components should
be more or less on the same scale (i.e., have approximately the same MTTF).


# Bootstrapping Statistics of the MLE and Likleihood Mode

To estimate various characteiristcs, such as the bias, BCa confidence intervals,
etc, then bootstraping may be used. The `likelihood.md.series.systems` package
relies upon the bootstrapping functionality in `boot` and `likelihood.model`,
but provide special functions and methods that are particular to the masked data
series system context.




# Documentation
For more detailed information on how to use this package and what each function does, please refer to the package documentation. The series_system function and its parameters are described in detail there.

# Contributing
Contributions are welcome! Please open an issue or submit a pull request on GitHub if you find any bugs or if you'd like to suggest improvements.

