---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# Series Systems: Estimating Lifetime Distributions of Unobserved Components with Masked Data

This R package, `series_system_estimation_masked_data`, provides a set of functions for
generating MLEs for the lifetime parameters of series systems and other related
characteristics from *masked data*.

Masked data comes in a variety of forms:

1. The system lifetime can be masked in three related ways.
   First, right censoring occurs when the system under observation
   is only known to have survived for some minimum length of time.
   Second, left censoring occurs when the system under observation
   is only know to have survived for some maximum length of time.
   Finally, interval censoring occurs when the system under observation
   is only known to have survived between some minimum and maximum
   length of time.

   In the unmasked situation, we know precisely how long the system
   under observation survived.

2. Regardless of how the series system lifetime is masked, the
   lifetime of the components may be masked in any of the ways
   described in item (1). There is, additionally, another kind of
   masking we would like to consider. What if we do not observe
   any of the component lifetimes, and instead are only given
   a (potentially masked) series system lifetime, and a *candidate
   set* of component indexes which plausibly contains the failed
   component index.

   For a series system of $m$ components, the candidate sets are
   subsets of $\{1,\ldots,m\}$, excluding the empty set in cases
   where we observe a series system failure.

This R package is mostly concerned with masked data that takes the
form of a (potentially masked) series system lifetime and a set
of candidate components. (If the series system lifetime is right
censored, then the candidate set is naturally empty since no
component has failed yet.)

<!-- badges: start -->
<!-- badges: end -->


## Installation

You can install the development version of `series_system_estimation_masked_data`
like so:

``` r
devtools::install_github("queelius/series_system_estimation_masked_data")
```

## Example

This is a basic example which shows you how to solve a common problem:

```{r example}
#library(series_system_estimation_masked_data)

## basic example code
```


## TODO

1. might be interesting to predict system lifetime based on candidate covariates.

2. might be interesting to predict node lifetimes based on candidate covariates and system lifetime

3. extract sampler and other related stuff, like random variables (not series variations)
   into a algebraic_random_elements R package.

4. compare info/cov with bootstrap version


# The formula interface for masked data model fitting

The function to fit a masked data model take two arguments,
a formula expressing the structure of the model and a source of data;
it then returns an object representing the fitted model consistent
with the arguments.

The `md` function fits a masked data model using maximum likelihood
estimation to estimate the parameters of the parameters of the latent
component lifetimes.

It takes two arguments, a formula expressing the structure of the model
and a source of data (a data frame, typically).

The object representing the fitted model is a child of `md_mle`.
`md_mle` inherits from `mle`, which defines an algebra over MLEs.
