---
title: "Series system with Weibull distributed component lifetimes"
output:
    rmarkdown::html_vignette:
        toc: true
vignette: >
  %\VignetteIndexEntry{Series system with exponentially distributed component lifetimes}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  echo = TRUE,
  comment = "#>")

library(algebraic.mle)
library(md.tools)
library(tidyverse)
library(devtools)
options(digits=4)
```


Introduction
============

The R package `series.system.estimation.masked.data` is a framework for
estimating the parameters of latent component lifetimes from *masked data*
in a series system. The masked data comes from two sources:

(1) The system might be right-censored.
(2) The cause of failure of the failed component might be masked.


Statistical model
=================

In this study, the system is a series system with $m$
components. The true DGP for the system lifetime is in the
Weibull series system family, i.e., the component lifetimes are
Weibull and independently distributed and we denote the
true parameter value by $\theta$.

The principle object of study is $\theta$, which in the case of
the Weibull series system family consists of $2 m$ parameters:
$$
  \theta = (k_1, \lambda_1, \ldots, k_m, \lambda_m)'.
$$

We are interested in estimating the $\theta$ from masked data.
The masking comes in two independent forms:

- Censored system failure times, e.g., right-censoring. The system failure
time is the minimum of the component lifetimes, and it is right-censored
if the system failure time does not occur during the observation period,
$$
    T_i = \min\{\tau_i, T_{i 1}, \ldots, T_{i m}\},
$$
where $\tau_i$ is the right-censoring time for the $i$\textsuperscript{th}
observation and $T_{i 1},\ldots,T_{i m}$ are the component lifetimes
for the $i$th system.

- The cause of failure, the failed component, is masked. This masking
comes in the form of a candidate set $\mathcal{C}_i$ that, on average,
conveys information about the component cause of failure.

The candidate set $\mathcal{C}_i$ is a random variable that is a subset of
$\{1,\ldots,m\}$. The true DGP for the candidate set model has a general form
that may be denoted by
$$
    \Pr\{\mathcal{C}_i=c_i|T_1=j,\ldots,T_m,\theta,\text{other factors}\}.
$$

This is a pretty complicated looking model, and we are not even interested
in the DGP for candidate sets, except to the extent that it affects the
sampling distribution of the MLE for $\theta$.

In theory, given some candidate set model, we could construct a joint likelihood
function for the full model and jointly estimate the parameters of both the
candidate set model and $\theta$. In practice, however, this could be a very
challenging task unless we make some simplifying assumptions about the DGP for
candidate sets.

## Candidate set models
In every model we consider, we assume that the candidate set $\mathcal{C}_i$
is only a function of the component lifetimes $T_{i 1},\ldots,T_{i m}$, $\theta$,
and the right-censoring time $\tau_i$. That is, the candidate set $\mathcal{C}_i$
is independent of any other factors (or held constant for the duration of the
experiment), like ambient temperature, and these factors also have a neglible
effect on the series system lifetime and thus we can ignore them.

### Reduced likelihood model
In the Bernoulli candidate set model, we make the following assumptions about
how candidate sets are generated:

  - $C_1$: The index of the failed component is in the candidate set,
  i.e., $\Pr\{K_i \in \mathcal{C}_i\} = 1$, where
  $K_i = \arg\min_j \{ T_{i j} : j = 1,\ldots,m\}$.

  - $C_2$: The probability of $C_i$ given $K_i$ and $T_i$ is equally
  probable when the failed component varies over the components in the candidate
  set, i.e., $\Pr\{\mathcal{C}_i=c_i|K_i=j,T_i=t_i,\theta\} = \Pr\{C_i=c_i|K_i=j',T_i=t_i\}$ for
  any $j,j' \in c_i$.
    
  - $C_3$: The masking probabilities are conditionally independent of $\theta$
  given $K_i$ and $T_i$, i.e., $\Pr\{\mathcal{C}_i=c_i|K_i=j,T_i=t_i\}$ is not a
  function of $\theta$.

Using these simplifying assumptions, we can arrive at a reduced likelihood
function that only depends on $\theta$ and the observed data and as long as our
candidate set satisfies conditions $C_1$, $C_2$, and $C_3$, our reduced
likelihood function obtains the same MLEs as the full likelihood function.

We see that
$$
    \Pr\{\mathcal{C}_i=c_i,|K_i=j,T_i=t_i\} = g(c_i,t_i),
$$
since the probability cannot depend on $j$ by condition $C_2$ and cannot depend
on $\theta$ by condition $C_3$. Thus, we can write the likelihood function as
$$
    L(\theta) = \prod_{i=1}^n f_{T_i}(t_i|\theta) g(c_i,t_i).
$$

We show that $g(c_i,t_i)$ is proportional to
$$
    g(c_i,t_i) \propto \sum_{j \in c_i} f_j(t_i|\theta_j) \prod_{l=j,l \neq j}^m R_l(t_i|\theta_l),
$$
and thus 

Note, however, that different ways in which the conditions are met will yield
MLEs with different sampling distributions, e.g., more or less efficient estimators.






### Bernoulli candidate set model #1
This is a special case of the reduced likelihood model.
In this model, we satisfy conditions $C_1$, $C_2$, and $C_3$, but we include 
each of the non-failed components with a fixed probability $p$, $0 < p < 1$.

In the simplest case, $p = 0.5$, and candidate set $c_i$ has a probability
given by
$$
\Pr\{\mathcal{C}_i=c_i|K_i=j,T_i=t_i\} =
\begin{cases}
        (1/2)^{m-1} & \text{if $j \in c_i$ and $c_i \subseteq \{1,\ldots,m\}$} \\
        0 & \text{if $j \notin c_i$}.
\end{cases}
$$
\begin{proof}
Since there are $m-1$ non-failed components (the failed component $j$ is 
in $c_i$ with probability $1$), there are $2^(m-1)$ possible
candidate sets (the size of the power set of the non-failed component indexes).
Each of these candidate sets has equal probability of occurring, and thus
the probability of any particular candidate set is $1/2^(m-1)$.
\end{proof}


### Bernoulli candidate set model #2

Now, we remove condition $C_2$. We still assume conditions $C_1$ and $C_3$,
but we allow $C_i$ to depend on the failed component $K_i$, i.e.,
$$
    \Pr\{\mathcal{C}_i=c_i|K_i=j,T_i=t_i,\theta\} \neq \Pr\{C_i=c_i|K_i=j',T_i=t_i\}
$$
for $j,j' \in c_i$.

In this case, we can write the likelihood function as
$$
    L(\theta) = \prod_{i=1}^n f_{T_i}(t_i|\theta) \prod_{j=1}^m \Pr\{K_i=j|T_i=t_i\} \prod_{c_i \in \mathcal{C}_i} g(c_i,t_i,j).
$$


Weibull series system
=====================

Suppose a Weibull series system with $m=5$ components is parameterized by
the following R code:

```{r}
theta <- c(1,    1, # component 1's shape and scale parameters
           1.1,  1, # component 2's shape and scale parameters
           0.95, 1, # component 3's shape and scale parameters
           1.15, 1, # component 4's shape and scale parameters
           1.1,  1) # component 5's shape and scale parameters
m <- 5
```

So, in our study, $\theta = (`r theta`)'$.
The component assigned to index $2$ has a Weibull distributed
lifetime parameterized by  $\theta_2 = (`r theta[3]` `r theta[4]`)$.

Let's simulate generating the lifetimes of the $m = `r m`$ components for this series
system:
```{r}
set.seed(7231) # set seed for reproducibility
n <- 75
comp_times <- matrix(nrow=n,ncol=m)
for (j in 1:m)
    comp_times[,j] <- rweibull(n,theta[2*j-1],theta[2*j])
comp_times <- md_encode_matrix(comp_times,"t")
print(comp_times,n=4)
```

Next, we use the function `md_series_lifetime_right_censoring` to decorate the
masked data with the right-censor-censoring time $\tau_1 = \tau_2 = \cdots = tau_n = 3$:
```{r}
tau <- 3
data <- comp_times %>% md_series_lifetime_right_censoring(tau)
print(data,n=4,drop_latent=TRUE)
```

## Masked component cause of failure (candidate sets)
We simulate candidate sets using the Bernoulli candidate model with an
appropriate set of parameters to satisfy conditions $C_1$, $C_2$, and $C_3$:
```{r warning=F, message=F}
p <- .3
data <- data %>% md_bernoulli_cand_C1_C2_C3(p)
print(data[,paste0("q",1:m)],n=4)
```

Now, to generate candidate sets, we sample from these probabilities:
```{r}
data <- data %>% md_cand_sampler()
print(md_boolean_matrix_to_charsets(data,drop_set=TRUE),drop_latent=TRUE,n=6)
```

We see that after dropping latent (unobserved) columns, we only have the right
censoring time, right censoring indicator, and the candidate sets. (Note that
this time we showed the candidate sets in a more friendly way using
`md_boolean_matrix_to_charsets`.)

## Log-likelihood of $\theta$ given masked data

